---
title: 'Lab 5: Dplyr, Pipes, and More'
author: "Statistical Computing, 36-350"
output: pdf_document
---

```{r, include=FALSE}
# A hook to wrap output based on a linewidth chunk option
# From https://github.com/yihui/knitr-examples/blob/master/077-wrap-output.Rmd
library(knitr)
hook_output = knit_hooks$get('output')
knit_hooks$set(output = function(x, options) {
  # this hook is used only when the linewidth option is not NULL
  if (!is.null(n <- options$linewidth)) {
    x = knitr:::split_lines(x)
    # any lines wider than n should be wrapped
    if (any(nchar(x) > n)) x = strwrap(x, width = n)
    x = paste(x, collapse = '\n')
  }
  hook_output(x, options)
})
opts_chunk$set(cache=TRUE, autodep=TRUE, cache.comments=TRUE, linewidth=79)
```

Name: Rufus Petrie

**This week's agenda**: mastering the pipe operator `%>%`, practicing `dplyr` verbs, and pivoting using `tidyr`.

Loading the tidyverse
===

Now we'll load the tidyverse suite of packages. (You should already have `tidyverse` installed from the last lab; but if for some reason you still need to install again, then you can just look back at the last lab's instructions.) This gives us access to the pipe operator `%>%` as well as the `dplyr` and `tidyr` packages needed to complete this lab.

```{r}
library(tidyverse)
```

Q1. Pipes to base R
===

For each of the following code blocks, which are written with pipes, write equivalent code in base R (to do the same thing).

- **1a.** 

```{r}
letters %>%
  toupper %>%
  paste(collapse="+") 
```

```{r q1a}
paste(toupper(letters), collapse = "+")
```

- **1b.** 

```{r}
"     Ceci n'est pas une pipe     " %>% 
  gsub("une", "un", .) %>%
  trimws
```

```{r q1b}
trimws(gsub("une", "un", "     Ceci n'est pas une pipe     "))
```

- **1c.**

```{r}
rnorm(1000) %>% 
  hist(breaks=30, main="N(0,1) draws", col="pink", prob=TRUE) 
```

```{r q1c}
hist(rnorm(1000), breaks=30, main="N(0,1) draws", col="pink", prob=TRUE) 
```

- **1d.** 

```{r}
rnorm(1000) %>% 
  hist(breaks=30, plot=FALSE) %>%
  `[[`("density") %>%
  max
```

```{r q1d}
max('[['((hist(rnorm(1000), breaks = 30, plot = FALSE)), "density"))
```

Q2. Base R to pipes
===

For each of the following code blocks, which are written in base R, write equivalent code with pipes (to do the same thing).

- **2a.** Hint: you'll have to use the dot `.`, as seen above in Q1b, or in the lecture notes.

```{r}
paste("Your grade is", sample(c("A","B","C","D","R"), size=1))
```

```{r q2a}
c("A","B","C","D","R") %>% 
  sample(., size = 1) %>% 
  paste("Your grade is", .)
```

- **2b.** Hint: you can use the dot `.` again, in order to index `state.name` directly in the last pipe command.

```{r}
state.name[which.max(state.x77[,"Illiteracy"])] 
```

```{r 2qb}
as.data.frame(state.x77) %>% 
  select(Illiteracy) %>% 
  arrange(desc(Illiteracy)) %>% 
  head(1)
```

- **2c.** Note: `str.url` is defined for use in this and the next question; you can simply refer to it in your solution code (it is not part of the code you have to convert to pipes).

```{r}
str.url = "http://www.stat.cmu.edu/~ryantibs/statcomp/data/king.txt"

lines = readLines(str.url)
text = paste(lines, collapse=" ")
words = strsplit(text, split="[[:space:]]|[[:punct:]]")[[1]]
wordtab = table(words)
wordtab = sort(wordtab, decreasing=TRUE)
head(wordtab, 10)
```

```{r q2c}
readLines(str.url) %>% 
  paste(., collapse = " ") %>% 
  strsplit(., split="[[:space:]]|[[:punct:]]") %>% 
  unlist() %>%
  table() %>% 
  sort(., decreasing = TRUE) %>%
  head(., 10)
```

- **2d.** Hint: the only difference between this and the last part is the line `words = words[words != ""]`. This is a bit tricky line to do with pipes: use the dot `.`, once more, and manipulate it as if were a variable name.

```{r}
lines = readLines(str.url)
text = paste(lines, collapse=" ")
words = strsplit(text, split="[[:space:]]|[[:punct:]]")[[1]]
words = words[words != ""]
wordtab = table(words)
wordtab = sort(wordtab, decreasing=TRUE)
head(wordtab, 10)
```

```{r q2d}
readLines(str.url) %>% 
  paste(., collapse = " ") %>% 
  strsplit(text, split="[[:space:]]|[[:punct:]]") %>%
  unlist %>% 
  .[. != ""] %>% 
  table() %>% 
  sort(., decreasing = TRUE) %>% 
  head(., 10)
```

Prostate cancer data set
===

Below we read in the prostate cancer data set, as visited in previous labs. 

```{r}
pros.df = 
  read.table("http://www.stat.cmu.edu/~ryantibs/statcomp/data/pros.dat")
```

Q3. Practice with `dplyr` verbs
===

In the following, use pipes and `dplyr` verbs to answer questions on `pros.df`.

- **3a.** Among the men whose `lcp` value is equal to the minimum value (across the entire data set), report the range (min and max) of `lpsa`. 

```{r q3a}
pros.df %>% 
  filter(lcp == min(lcp)) %>% 
  filter(lpsa == max(lpsa) | lpsa == min(lpsa)) %>% 
  select(lpsa)
```

- **3b.** Order the rows by decreasing `age`, then display the rows from men who are older than 70 and without SVI. 

```{r q3b}
pros.df %>% 
  arrange(desc(age)) %>% 
  filter(age > 70 & svi == 0)
```

- **3c.** Order the rows by decreasing `age`, then decreasing `lpsa` score, and display the rows from men who are older than 70 and without SVI, but only the `age`, `lpsa`, `lcavol`, and `lweight` columns. Hint: `arrange()` can take two arguments, and the order you pass in them specifies the priority. 

```{r q3c}
pros.df %>% 
  arrange(desc(age), desc(lpsa)) %>% 
  filter(age > 70 & svi == 0) %>% 
  select(age, lpsa, lcavol, lweight)
```

- **3d.** We're going to resolve Q2c from Lab 3 using the tidyverse. Using `purrr` and `dplyr`, perform t-tests for each variable in the data set, between SVI and non-SVI groups. To be precise, you will perform a t-test for each column excluding the SVI variable itself, by running the function `t.test.by.ind()` below (which is just as in Q2c in Lab 3). Print the returned t-test objects out to the console.

```{r}
t.test.by.ind = function(x, ind) {
  stopifnot(all(ind %in% c(0, 1)))
  return(t.test(x[ind == 0], x[ind == 1]))
} 
```

```{r q3d}
pros.df %>% 
  select(-c("svi")) %>%
  lapply(., t.test.by.ind, ind = pros.df[,5])
```

- **3e.** Extend your code from the last part (append just one more line of code, glued together by a pipe) to extract the p-values from each of the returned t-test objects, and print them out to the console.

```{r q3e}
pros.df %>% 
  select(-c("svi")) %>%
  lapply(., t.test.by.ind, ind = pros.df[,5]) %>% 
  sapply(., '[[', 'p.value')
```

Fastest 100m sprint times
===

Below, we read in two data sets of the 1000 fastest times ever recorded for the 100m sprint, in men's and women's track., as seen in the last lab. 

```{r}
sprint.m.df = read.table(
  file="http://www.stat.cmu.edu/~ryantibs/statcomp/data/sprint.m.txt", 
  sep="\t", quote="", header=TRUE)
sprint.w.df = read.table(
  file="http://www.stat.cmu.edu/~ryantibs/statcomp/data/sprint.w.txt", 
  sep="\t", quote="", header=TRUE)
```

Q4. More practice with `dplyr` verbs
===

In the following, use pipes and `dplyr` verbs to answer questions on `sprint.w.df`.

- **4a.** Order the rows by increasing `Wind` value, and then display only the women who ran at most 10.7 seconds. 

```{r q4a}
sprint.w.df %>% 
  arrange(Wind) %>% 
  filter(Time <= 10.7)
```

- **4b.** Order the rows by terms of increasing `Time`, then increasing `Wind`, and again display only the women who ran at most 10.7 seconds, but only the `Time`, `Wind`, `Name`, and `Date` columns. 

```{r q4b}
sprint.w.df %>% 
  arrange(Time, Wind) %>% 
  filter(Time <= 10.7) %>% 
  select(Time, Wind, Name, Date)
```

- **4c.** Plot the `Time` versus `Wind` columns, but only using data where `Wind` values that are nonpositive. Hint: note that for a data frame, `df` with columns `colX` and `colY`, you can use `plot(colY ~ colX, data=df)`, to plot `df$colY` (y-axis) versus `df$colX` (x-axis).

```{r q4c}
sprint.w.df %>% 
  filter(Wind <= 0) %>% 
  plot(Time ~ Wind, data = .)
```

- **4d.** Extend your code from the last part (append just two more lines of code, glued together by a pipe) to plot the single fastest `Time` per `Wind` value. (That is, your plot should be as in the last part, but among points that share the same x value, only the point with the lowest y value should be drawn.)

```{r q4d}
sprint.w.df %>% 
  filter(Wind <= 0) %>%
  group_by(Wind) %>%
  filter(Time == min(Time)) %>% 
  plot(Time ~ Wind, data = .)
```

Q5. Practice pivoting wider and longer
===

In the following, use pipes and `dplyr` and `tidyr` verbs to answer questions on `sprint.m.df`. In some parts, it might make more sense to use direct indexing, and that's perfectly fine.

- **5a.** Confirm that the `Time` column is stored as character data type. Why do you think this is? Convert the `Time` column to numeric. Hint: after converting to numeric, there will be `NA` values; look at the position of one such `NA` value and revisit the original `Time` column to see why it was stored as character type in the first place.

```{r q5a}
typeof(sprint.m.df$Time)
sprint.m.df <- sprint.m.df %>% 
  mutate_at("Time", as.numeric)
```
One such NA value had the time "9.98A", which means that the time variable could be a character because of an input error.

- **5b.** Define a reduced data frame `dat.reduced` as follows. For each athlete, and each city, keep the fastest of all times they recorded in this city. Then drop all rows with an `NA` value in the `Time` column Your new data frame `dat.reduced` should have 600 rows and 3 columns (`Name`, `City`, `Time`). Confirm that it has these dimensions, and display its first 10 rows. Hint: `drop_na()` in the `tidyr` package allows you to drop rows based on `NA` values.

```{r q5b}
dat.reduced <- sprint.m.df %>% 
  group_by(Name, City) %>%
  filter(rank(Time, ties.method ="first") == 1) %>%
  drop_na(., Time) %>% 
  select(Name, City, Time)
head(dat.reduced, 10)
```

- **5c.** The data frame `dat.reduced` is said to be in "long" format: it has observations on the rows, and variables (`Name`, `City`, `Time`) on the columns. Arrange the rows alphebtically by city; convert this data frame into "wide" format; and then order the rows so that they are alphabetical by sprinter name. Call the result `dat.wide`. To be clear, here the first column should be the athlete names, and the remaining columns should correspond to the cities. Confirm that your data frame has dimension 141 x 152 Do these dimensions make sense to you?

```{r q5c}
dat.wide <- dat.reduced %>% 
  arrange(City) %>% 
  pivot_wider(names_from = City,
              values_from = Time)
```

- **5d.** Not counting the names in the first column, how many non-`NA` values does `dat.wide` have? How could you have guessed this number ahead of time, directly from `dat.reduced` (before any pivoting at all)?

```{r q5d}
sum(is.na(dat.wide))
```
You could have surmised this number from 'dat.reduced' by taking the total number of (athletes x cities) and subtracting the amount of times.

- **5e.** From `dat.wide`, look at the row for "Usain Bolt", and determine the city names that do not have `NA` values. These should be the cities in which he raced. Determine these cities directly from `dat.reduced`, and confirm that they match.

```{r q5e}
dat.wide[dat.wide$Name == "Usain Bolt",]
dat.reduced[dat.reduced$Name == "Usain Bolt",]
```

- **5f.** Convert `dat.wide` back into "long" format, and call the result `dat.long`. Remove rows that have `NA` values (hint: you can do this by setting `values_drop_na = TRUE` in the call to the pivoting function), and order the rows alphabetically by athlete and city name. Once you've done this, `dat.long` should have matching entries to `dat.reduced`; confirm that this is the case.

```{r q5f}
dat.long <- dat.wide %>% 
  pivot_longer(names_to = "City",
               values_to = "Time",
               cols = 2:152,
               values_drop_na = TRUE) %>% 
  arrange(Name, City)
dat.reduced <- dat.reduced %>% 
  arrange(Name, City)
head(dat.long)
head(dat.reduced)
```
The dataframes have the same entries.